Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-19 paper.tex
...

\abstract{Abstract.

@@

\noindent \textbf{Keywords:} Radiosity, global illumination, constant time.
}

@@

Pingus is an open-source clone of Lemmings%
\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
a puzzle-platformer video game (Figure~\ref{fig.pingus}).      
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit%
\footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

According to Tim Sweeney (of Unreal Engine fame), about half the complexity in
game development resides in \emph{simulation} (aka \emph{game logic}), but
which accounts for only 10\% of the CPU budget~\cite{games.sweeney}
(Figure~\ref{fig.sweeney}).
The game logic ``models the state of the game world as interacting objects
evolve over time''.
The high development costs contrasting with the low impact on performance
appeals for alternatives with productivity in mind, especially considering that
it is the game logic that varies the most between projects.
Sweeney states that ``will gladly sacrifice 10\% of our performance for 10\%
higher productivity''.

@@

Object-oriented games use the \emph{observer pattern}~\cite{games.patterns}
in the game logic to handle events from the environment (e.g., key presses and
timers) and also as a notification mechanism between game entities.
%
The observers are short-lived callbacks which must execute as fast as possible
and in real time to keep the game reactive to incoming events.
%
For this reason, callbacks cannot contain long-lasting locals and loops, which
are elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \code{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{http://tirania.org/blog/archive/2013/Aug-15.html}}%
\footnote{``Escape from Callback Hell'':
\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

\CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that aims to offer
a concurrent and expressive alternative to C/C++ with the characteristics that
follow:
%
\begin{itemize}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control mechanisms, such as
      \code{await} (to suspend a line of execution), and \code{par} (to combine
      multiple lines of execution).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      line of execution, i.e., there's no implicit preemption or real
      parallelism.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and the language requires no garbage
collection.

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
In Pingus, the game logic also accounts for almost half the size of the
codebase (18.173 from 39.362 LoCs, or 46\%).

rewriting Pingus to \CEU is to

Contributions:
    - present SSRP as an effective alternative for game logic dev
    - qualitative analysis
        - patterns
            - solutions

@@

In Pingus, the game logic also accounts for almost half the size of the
codebase (18.173 from 39.362 LoCs, or 46\%).

We rewrote 126 of the 272 files (46\%) which account for 9.186 of the 18.173
LoCs (51\%) comprising the game logic of Pingus [[![X]][cpp\_uncompressed]].
Half of the game logic relates to non-reactive code, such as configurations and
options, saved games and serialization, maps and levels descriptions, string
formatting, collision detection, graph algorithms, etc.
This part remains unchanged and relies on the seamless integration between \CEU
and C/C++ to remain usable.
From the 9.186 touched LoCs, we removed all headers, declarations, trivial
getters \& setters, and other innocuous statements, resulting in 70
implementation files with 4.135 dense LoCs originally written in C++
[[![X]][cpp\_compressed]].
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs
[[![X]][ceu\_compressed]].
Figure~\ref{tab.tree} summarizes the condensed codebase in the two
implementations:

\footnote{ Game logic code: \url{https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP} }.
%[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
%[ceu_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL

This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
<!--, helping on encapsulation and cohesion.-->
Nonetheless, the lowest ratio numbers above correlate to the parts of the game
logic that we consider more susceptible to structured reactive programming.
For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
animations that are affected by timers, game rules, and user interaction.
In contrast, the \emph{Option screen} (\emph{ratio 0.97}) is a simple UI grid
with trivial mouse interactions.

@@

We selected 6 representative game behaviors and describe their
implementations in C++ and \CEU.
We also categorized these examples in 5 abstract C++ control-flow patterns that
likely apply to other games:

\begin{enumerate}
\item \emph{Finite State Machines}:
    State machines describe the behavior of entities by mapping event
    occurrences to transitions between states that trigger appropriate actions.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next.
\item \emph{Dispatching Hierarchies}:
    Entities typically form a dispatching hierarchy in which a container that
    receives a stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities typically form a lifespan hierarchy in which a terminating
    container entity automatically destroys its managed children.
\item \emph{Signaling Mechanisms}:
    Entities often need to communicate explicitly through signaling mechanisms,
    especially if there is no hierarchy relationship between them.
\end{enumerate}

@@

The \emph{mouse grab option} restricts the mouse movement to the game window
boundaries%
\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
In addition, the \emph{Options menu} has a check box to toggle the
\emph{mouse grab option} with mouse clicks while still responding to
\emph{Ctrl-G} presses.
%
Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.

@@

>>>>

\abstract{TODO.

@@

\noindent \textbf{Keywords:} TODO, TODO, TODO.
}

@@

Pingus is an open-source puzzle-platform video game based on Lemmings.
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit (Figure~\ref{fig.pingus}).
% \footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

According to Tim Sweeney (of Unreal Engine fame), about half the complexity in
game development resides in \emph{simulation} (aka \emph{game logic}), but
which accounts for only 10\% of the CPU budget~\cite{games.sweeney}
(Figure~\ref{fig.sweeney}).
%The game logic ``models the state of the game world as interacting objects
%evolve over time''.
The high development costs contrasting with the low impact on performance
appeals for alternatives with productivity in mind, especially considering that
it is the game logic that varies the most between projects.
Sweeney states that ``will gladly sacrifice 10\% of our performance for 10\%
higher productivity''.

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot contain long-lasting locals and loops, which
are elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \code{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{http://tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

\CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that offers a
concurrent and expressive alternative to C/C++ with the characteristics that
follow:
%
\begin{itemize}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control mechanisms, such as
      \code{await} (to suspend a line of execution), and \code{par} (to combine
      multiple lines of execution).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      line of execution, i.e., there's no implicit preemption or real
      parallelism.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection.

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting the game logic of Pingus from C++ to \CEU
with the contributions that follow:
%
\begin{itemize}
\item Identification of five recurrent control-flow patterns in the game logic
      that likely apply to other games.
\item Idiomatic solutions in \CEU for six selected behaviors in the game logic.
\item An in-depth qualitative analysis of the proposed solutions in comparison
      to C++.
\end{itemize}

@@

In Pingus, the game logic also accounts for almost half the size of the
codebase: 18.173 from 39.362 LoCs (46\%) spread in 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
configurations and options, saved games and serialization, maps and levels
descriptions, string formatting, collision detection, graph algorithms, etc.
This part remains unchanged and relies on the seamless integration between \CEU
and C/C++~\cite{ceu.sensys13}.
%
Therefore, we rewrote 9.186 LoCs spread in 126 files%
\footnote{Game logic codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/cpp/}}.
%
In order to only consider effective code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting in 70 implementation files with 4.135 dense LoCs
originally written in C++%
\footnote{``Effective'' C++ codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/all/}}.
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
\footnote{``Effective'' \CEU codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/all/}}.
%
Figure~\ref{tab.tree} summarizes the effective codebase in the two
implementations.

This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%<!--, helping on encapsulation and cohesion.-->
Nonetheless, the lines with lower ratio numbers above correlate to the parts of
the game logic that we consider more susceptible to structured reactive
programming.
For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
animations that are affected by timers, game rules, and user interaction.
In contrast, the \emph{Option screen} (\emph{ratio 0.97}) is a simple UI grid
with trivial mouse interactions.

@@

We selected 6 representative game behaviors and describe their
implementations in C++ and \CEU.
We also categorize these behaviors in 5 abstract control-flow patterns that
likely apply to other games:

\begin{enumerate}
\item \emph{Finite State Machines}:
    Event occurrences map to transitions between states that trigger
    appropriate actions comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next in the game.
\item \emph{Dispatching Hierarchies}:
    Entities typically form a dispatching hierarchy in which a container that
    receives a stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities typically form a lifespan hierarchy in which a terminating
    container entity automatically destroys its managed children.
\item \emph{Signaling Mechanisms}:
    Entities often need to communicate explicitly through signaling mechanisms,
    especially if there is no hierarchy relationship between them.
\end{enumerate}

@@

The \emph{mouse grab option} restricts the mouse movement to the game window
boundaries%
\footnote{Mouse grab animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#6} }.
The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
In addition, the \emph{Options menu} has a check box to toggle the
\emph{mouse grab option} with mouse clicks while still responding to
\emph{Ctrl-G} presses.
%
Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.

@@

TODO: non reactive, C++ integration
